
\part{Annotation Exchange Format for Images}
\section{Overview}

\p{Image annotation and segmentation is an important 
analytic process in many scientific, technical, and 
commercial fields.  Nonetheless, there are few standard 
formats for describing and representing image annotations, and 
those which do exist tend to be used in specific, relatively 
narrow contexts.\footnote{Current formats include \sAIM{}  
(Annotation and Image Markup), \sCVAT{} \sXML{} (\sCVAT{} is 
the Computer Vision Annotation Tool), \sDICOMSR{} (Digital 
Imaging and Communications in Medicine Structured Reporting), 
\sPASCAL{} \sVOC{} \sXML{} (Pattern Analysis, Statistical Modeling and 
Computational Learning Visual Object Classes), 
and \sCOCO{} \sJSON{} (Common Objects in Context).}  This is 
not a new observation; Daniel L. Rubin \textit{et. al.}, 
in 2007, note that: 

\begin{displayquote}
Images contain implicit knowledge about anatomy and abnormal structure that 
is deduced by the viewer of the pixel data, but this knowledge is generally 
not recorded in a structured manner nor directly linked to the image.  
[Moreover,] the \textit{terminology} and \textit{syntax} 
for describing images and what they 
contain varies, with no widely-adopted standards, resulting in limited interoperability.  The contents of medical images are most frequently 
described and stored in free-text in an unstructured manner, limiting the 
ability of computers to analyze and access this information.  There are no  standard terminologies specifically for describing medical image contents 
--- the imaging observations, the anatomy, and the pathology.  [N]o 
comprehensive standard appropriate to medical imaging has yet been developed.
A final challenge for medical imaging is that the particular information one 
wants to describe and annotate in medical images depends on the \textit{context} 
--- different types of images can be obtained for different purposes, and the  types of annotations that should be created (the \q{annotation requirements} 
for images) depends on that context.  For example, in images of the abdomen of 
a cancer patient (the context is \q{cancer} and \q{abdominal region}), we 
would want  annotations to describe the liver (an  organ in the abdominal region), and if there is a cancer in the liver, then there should be a description 
of the margins of the cancer (the appearance of the cancer on the image).
(\bhref{http://cedarweb.vsp.ucar.edu/wiki/images/d/d9/R_19.pdf}, pp. 1-2). 
\end{displayquote}

These challenges inspired \AIM{} (the \q{Annotation and Image Markup} 
project), which \q{provides a solution to the ... imaging challenges 
[of]: No agreed upon syntax for annotation and markup;  
No agreed upon semantics to describe annotations; No standard format 
... for annotations and markup} (\bhref{https://wiki.nci.nih.gov/display/AIM/Annotation+and+Image+Markup+-+AIM}).  However, \AIM{} itself has not 
been widely adopted outside the specific field of cancer research 
and cancer-oriented image repositories.}

\p{One obstacle to formalizing image-annotation data is that 
annotations have a kind of intermediate status, neither intrinsic 
parts of an image nor merely visual cues supporting the presentation 
of the image within image-viewing software.  Many applications 
exist which allow markup or comments to be introduced with 
respect to an image.  From the application's point of view, 
these annotations are part of the application display, 
not part of the image --- analogous to editing comments that 
might be added to a text document by a word processor or 
\PDF{} viewer, which are records of user actions, not 
intrinsic to the document itself.  Indeed, one 
mechanism for recording image annotations in \DICOM{} 
(the \q{Digital Imaging and Communications in Medicine} 
format) is via \q{presentation state.}  The presentation 
state includes all details about how the image currently 
appears to \DICOM{} workstation users, such as  
Radiologists, which could include markings they have made 
to indicate diagnostically significant image regions or 
features.  Insofar as image-annotations are 
considered to be artifacts of image-viewing software, 
rather than significant data structures in their 
own right, there is less motivation for imaging applications 
to support canonical annotation standards.}

\p{Nevertheless, in many scientific and technical 
areas image annotations \textit{are} significant; 
they are intrinsic to the scientific value of a 
given image as an object of research or observation.  
Image regions, segments, and features have a 
semantic meaning outside the contexts of the 
applications that are used to view the 
corresponding images, which is why it is important 
to develop cross-application standards for 
describing and affixing data to image annotations.}

\p{It is also important for image-annotation 
models to be broadly applicable and multi-disciplinary.  
While image analysis serves different goals in 
different contexts (e.g., segmentation of microscope images 
to detect cancer cells serves different ends than 
segmentation of camera snapshots to study traffic patterns), 
there is always a possibility of analytic techniques 
developed in one subject area to be applicable for 
other image-processing problems, even if the practical 
outcomes desired of the analyses are very different.  
Furthermore, certain computational domains are 
similar enough to image analysis to warrant inclusion 
in a general-purpose image-annotation framework, 
even if the underlying data is not \q{images} in the 
conventional sense (not, for instance, captured 
via photographs or microscopy).  For example, 
\PDF{} document views, Flow Cytometry (\FCM{}) data plots, 
and geospatial maps subject to Geographic Information 
Systems (\GIS{}) annotations may all be considered 
images --- by virtue of a semantic significance attributed 
to color and to geometric primitives as a way 
of characterizing phenomena observed or modeled through 
their data --- even though such resources are 
not acquired by ordinary \q{image-producing} devices.  
The \q{Pantheon} project, characterized as a
\q{platform dedicated to knowledge engineering for the 
development of image processing 
applications,}\footnote{See \bhref{https://hal.archives-ouvertes.fr/hal-00260065/document}.}
offers one of the few attempts in imaging literature to 
rigorously define \q{imaging} and \q{image processing} in the 
first place.  Pantheon (via its \Pandore{} component) 
also includes an image processing 
\textit{objectives} ontology, which is partially incorporated 
into \AXFI{}, so Pantheon's articulation of the 
general domain of imaging \textit{per se} applies to \AXFI{} 
as well.  The problem of defining \q{images} as such, 
and therefore delineating the scope of image annotation, is 
addressed below (page~\lpageref{ph:acquisition}).  In brief, \AXFI{} 
considers the realm of imaging to be more general 
than just graphics obtained by a direct recording 
of the optics of some physical scene via cameras, 
microscopes, or telescopes.  That is to say, 
the image acquisition process is not necessarily 
one where data is generated by an instrument which 
produces a digital artifact by absorbing light, 
so that geometric and chromatic properties of the 
image are wholly due to the functioning of 
the acquisition device.}

\p{Systematically identifying the scope of \q{image annotation} 
is important for \AXFI{} because doing so clarifies 
the sorts of domains whose semantics could 
reasonably be incorporated into \AXFI{}, as 
well as the sorts of applications which would 
be reasonable candidates for supporting 
\AXFI{} annotations (i.e., the capability 
to parse \AXFI{} data and represent it 
\visavis{} the relevant images).  For 
example, if immunoflourescent 
Flow Cytometry (\FCM{}) data plots are 
classified as images, then the numerical 
properties of the \q{channel} axis, with 
notions of \q{decades} and a \q{log/linear} 
distinction, become relevant to the \AXFI{} 
vocabulary for representing spatial 
dimensions and magnitudes.  In general, 
\AXFI{} uses paradigms and terminology 
from \q{Conceptual Space Theory} as 
part of the process of formalizing 
geometric and dimensional 
notions.\footnote{See \bhref{http://idwebhost-202-147.ethz.ch/Publications/RefConferences/ICSC_2009_AdamsRaubal_Camera-FINAL.pdf} or\\ 
\bhref{https://arxiv.org/pdf/1801.03929.pdf}.} }

\p{When defining the scope of \AXFI{}, it is also 
important to distinguish the \textit{data models} 
encapsulated by \AXFI{} resources from the 
file formats where \AXFI{} data is encoded.  
The choice of one file type or another 
to represent a data structure --- \XML{} versus
\JSON{}, for instance --- does not fundamentally 
affect the data thereby communicated.  
Therefore, it is important to formalize data 
models in such a way that numerous different 
serialization languages might actually be 
used to share/express the data.  However, 
in practice, format-specific standards, 
such as \XML{} Schema Definitions, 
are often used as the basis for 
formalizing and enforcing compliance 
with data models.  Therefore, \AXFI{} 
cannot be completely unconcerned with 
the structure and requirements of 
files which convey \AXFI{} data.  
This is particularly true because \AXFI{} 
seeks to incorporate the data models 
of other specifications, such as 
\AIM{} and \GatingML{}, which are formalized 
via \XML{} specifications.  Although \AXFI{} 
is not primarily \XML{}-based, in short, it 
intends to be (in the relevant contexts) 
compatible with \XML{} languages that 
rely on \XML{} schematization.  Further 
details on how \AXFI{} manages the relation 
to \XML{} and other serialization languages 
are provided below (page~\lpageref{sec:SFour}).}

\p{A further detail that should be clarified 
prior to expositing a formal outline of 
\AXFI{} is that of how image-annotations originate.  
Sometimes, of course, annotations are manually 
introduced on images by human users of image-viewing 
software.  On the other hand, automated 
image segmentation --- or similar algorithmic 
or \AI{}-driven image processing without human 
intervention --- yields partitions of images into 
regions, or identification of semantically 
important locations in an image, therefore generating 
annotations computationally.  In short, \AXFI{} 
should support both human-generated and computer-generated 
annotations.  This becomes complicated, however, insofar 
as image-processing may yield analyses which overlap 
with annotation objectives but may not intrinsically 
produce annotations in the conventional sense.  
For example, an algorithm to count the number of 
cars in a highway picture may rely on statistical 
analysis of some quantitative image feature --- 
such as \q{zero-crossings} --- without in fact 
producing determinate image segments.}

\p{It is important 
to remember that image processing operations do not 
always act directly on images themselves; sometimes 
algorithms are based instead on mathematical complexes 
derived from the image, but with their own quantitative 
properties.  For instance, color-valued pixels may 
be replaced by matrices measuring the gradient 
of some image-feature field in eight directions around 
each point (an example would be \q{Sobel kernels} applied 
to the image intensity function).  For a given \q{semantic} 
task --- that is, an image-processing objective whose 
end-result is not just image-related data but some 
empirical observation --- image segmentation, 
or other analyses yielding annotations, are a means 
to an end: one \textit{way} to count cars is to 
delineate the edges of distinct cars in distinct 
segments, and then count the number of segments 
which result.  However, statistical image-analysis 
may produce largely accurate results for such 
semantic tasks, given large image corpora (e.g., estimating 
traffic flows from highway cameras), without yielding 
artifacts such as human-visible segment representations.  
Or, in a different domain, \AI{}-powered analysis 
of \FCS{} (Flow Cytometry Standard) data could establish a largely 
accurate count of \q{events} (i.e., discrete \FCS{} measurements 
of light-scattering and/or flourescent properties of 
cellular-scale entities) without manual \q{gating} (referring 
to the conventional practice of scientists using geometric 
annotations of \FCS{} data-plots to isolate and 
thereby count different event-types).  An 
\AI{}-powered analysis of image features, or 
likewise of Flow Cytometry data, may yield calculations 
similar to those which for \textit{human} users 
are achieved via image segmentation, manual gating, 
and similar operations which clearly yield 
annotation data.  For \AXFI{}, the complication 
arises when these \AI{} mechanisms do not 
themselves yield results that would normally 
be considered annotations, but rather yield 
the desired empirical results for which 
the annotations would be a preliminary 
step --- e.g., an approximate count of 
the number of cars in a highway photo, 
without a precise segmentation of the 
image marking the cars' respective borders.}

\p{\lAXFI{} ultimate considers these  
\AI{}-related complications to be issues 
resolvable at the level of software, 
rather than standardized annotation models 
\textit{per se}.  An image annotation is, 
among other things, a visual (or viewable) 
record of some image-processing activity.  
If a radiologist manually clarifies a 
report to the effect that a given \CT{} 
shows a tumor by circling the area where 
the tumor is visible, he or she is using 
the image annotation to communicate to 
others the thought-process which motivated 
the diagnostic conclusion.  This is different 
than an \AI{}-driven processor which 
would automatically demarcate an image segment 
outlining the tumor and use geometric properties 
of that segment to derive a pathological 
finding.  In short, the data conveyed in an 
annotation --- an image segment, rendered 
precisely, or rendered indirectly via a circle or 
polygon around the segment --- may be 
\textit{intrinsic} to an image-processing 
operation: it may be data acquired \textit{at 
one stage} in an analytic workflow.  However, 
annotations may also be \textit{retroactive}; 
if a radiologist circles a tumor, he or she 
has completed (at least mentally) the 
image analysis, and is using the analysis 
to summarize what occurred in the course of the analysis.  
Therefore, any image-processing task can be associated 
with \textit{ex post facto} annotations which 
summarize the process even if they are not 
intrinsic to it.}

\p{To continue the example of counting cars from a 
highway photo, an \AI{}-powered observation could 
be retroactively \textit{justified} by providing 
an image segmentation where the number of 
car-segments matches the \AI{} count.  In lieu 
of precise segments, it may be simpler to 
provide location-points for the \q{geometric center} of 
each car, or the points furthest apart in the 
direction of each car's front-to-back --- these 
may be the statistical signals used for 
the car-counting process.  Analogously, facial 
recognition does not need to rely one segmenting 
out regions (eyes, nose, lips), but rather can be 
based on distances between individual points (such 
as the inner corners of each eye).  But in any 
case, depending on the analytic algorithm used, 
it is often possible to identify some spatial/geometric 
feature or object that can be visualized in 
the image context, and that summarizes or 
legitimizes the analytic operation.  This 
summarial data, then, can provide 
\textit{retrospective} annotations which allow 
human viewers to understand and review the 
algorithmic process.  In short, simply because 
image-processing tasks may not generate 
annotation data as part of their internal 
activity, it is still possible (and 
may be desirable) for the software 
operationalizing these tasks to 
implement annotation generators, where 
the resulting annotations document 
the operations for the scientific record 
and/or summarize them in \GUI{} objects 
for the benefit of human viewers.}

\p{In general, then, \AXFI{} distinguishes 
human-generated from computer-generated 
annotations, and moreover leaves open 
the possibility that some computer-generated 
annotations are \textit{retrospective}: that 
instead of being internal to an imaging 
computation they are indirectly produced 
subsequent to such a computation, for 
purposes of documentation and validation. 
Annotations which are not \textit{retrospective} 
are called \textit{internal}, as in, 
internal to a given image-processing workflow.}

\p{Related to the distinction between 
\textit{internal} and \textit{retrospective} 
annotations, \AXFI{} recognizes a contrast 
between \q{intrinsic} and \q{descriptive} 
annotation.  An example of an \textit{intrinsic}  
annotation might be an image segment, 
where calculating the boundary of the segment 
is intrinsic to a specific image-processing 
objective, whereas an \textit{descriptive} 
annotation might be an arrow \textit{pointing 
toward} that segment.  Here the descriptive 
annotation is introduced primarily for the 
benefit of human viewers.}

\p{The intrinsic/descriptive distinction 
is not always clear-cut.  Consider the 
following two cases: in one scenario, an 
image-segmentation routine precisely 
delineates a region of interest (e.g., an outline 
of a red bird against blue sky), 
notating the segmentation result via a 
two-color-depth transform of the original 
image.  Image-viewing software then 
shows the segment indirectly by encircling 
it, producing, in effect, a secondary 
annotation intended to call attention to 
the primary (segment) annotation.  Here, 
clearly, the segment itself 
(encoding by a separate two-toned image) 
is intrinsic to the original analysis, whereas 
the secondary annotation has a purely 
descriptive purpose.} 

\p{However, consider 
an alternate scenario where the 
original segmentation is done 
with less precision (this may be 
the case where there is a more 
muted color differential between foreground 
and background).  Imperfect segmentation 
may still be adequate for some semantic task 
(e.g., identifying a bird's species).  
An analysis might obtain a rough segmentation 
by marking certain points highlighted by an 
edge-detector, then protruding the convex 
hull of the region outward so as to be sure 
of encompassing the whole bird-segment within a 
polygon, albeit allowing some background pixels 
into the polygon as well.  This approximate 
segment is only an indirect representation of 
the region of interest (which would be the 
avian sub-image clearly outlined with no background 
included), but for analytic purposes the 
rough polygon might be a reasonable substitute 
for the finer-grained segment, analogous to 
how a cubic or quartic polynomial may be an 
adequate approximation to a more complex curve.  
Depending on how it is used, the approximate 
segment may therefore be considered merely 
a visual cue connoting the region of interest, 
or a significant region in its own right.  
This contrast would, most likely, depend 
on whether the approximation is used 
itself as a basis for further analysis, 
or instead is mostly a presentation device.  
This usage-context would therefore indicate 
whether an \q{imprecise} annotation 
(\AXFI{} formally uses the term 
\textit{approximative}) should be classified 
as \textit{intrinsic} or \textit{descriptive}.}

\p{All told then, annotations may be 
\textit{internal} or \textit{retrospective}, 
and \textit{intrinsic} or \textit{descriptive}.  
\lAXFI{} also contrast \textit{computer-generated} 
from \textit{human-generated} annotations 
(which are related to but not the same terms 
as \textit{manual} and \textit{automated}; see 
below, page~\lpageref{ph:man}).  
These distinctions are indepenedent of one another; 
retrospective annotations for instance could 
be either intrinsic or descriptive, and either 
computer-generated or human-generated.}

\p{With these preliminaries concerning the 
scope of \AXFI{} clarified, the 
following sections will (1) outline 
the kind of data encoded via \AXFI{} 
and (2) the relation between \AXFI{}, 
data types, and applications/libraries 
that may use \AXFI{}.  For purposes of exposition, the following 
specifications will discuss images mostly 
in two (spatial) dimensions; however, \AXFI{} does not 
preclude annotations applied to image-sources 
with other dimensional profiles, such as 
\ThreeD{} graphics, or movies (\TwoD{} plus time), 
or even audio files (consistent with \DICOMSR{}, 
which supports audio annotations because 
\DICOM{} is sometimes used to share biomedical 
acoustical data, such as ultrasound).}

%\p{}


